### 2021-02-25 rad-icfp-2021/intro.tex
some refs

The lack of purity of reverse-mode is known to make it significantly harder to optimise and find parrallelism. 
None of the current implementations of reverse-mode in a functional setting are pure \cite{}, and often complicated heuristics with no guarantees are used \cite{}.
It means that for every important non trivial operation, a hand-crafted reverse-derivative must be given to be efficient.
In other words functional implementations seem to stuggle to abstract away from imperative code.

@@

\begin{example}[First-order term]
On simple first-order terms, the output code looks like SSA code \cite{}.
Let's consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> (y1+x2*y4, y2+x1*y4, y3)> in!\\
        !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
        !w_2'(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

After some general partial evaluation that will be detailed further in the paper, we obtain     
        \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1 = 0 in!\\
                !let y2 = 0 in!\\
                !let y3 = 0 in!\\
                !let y4 = 0 in!\\
                !let y5 = 1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
        \end{center}  
For the expert viewer, this is equivalent to the SSA form of what the imperative reverse-mode differentiation of our initial term would be.\\
Of course, this term can be further optimised to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}

>>>>

The lack of purity of reverse-mode is known to make it significantly harder to optimise and find parrallelism. 
None of the current implementations of reverse-mode in a functional setting \cite{lantern_icfp,pearlmutter2008reverse,baydin2016diffsharp} are pure, and often complicated heuristics with no guarantees are used, e.g. \cite{xla}.
It means that for every important non trivial operation, a hand-crafted reverse-derivative must be given to be efficient.
In other words functional implementations seem to stuggle to abstract away from imperative code.

@@

\begin{example}[First-order term]
On simple first-order terms, the output code looks like SSA code \cite{cytron1989efficient}.
Let's consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> (y1+x2*y4, y2+x1*y4, y3)> in!\\
        !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
        !w_2'(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

After some general partial evaluation that will be detailed further in the paper, we obtain     
        \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1 = 0 in!\\
                !let y2 = 0 in!\\
                !let y3 = 0 in!\\
                !let y4 = 0 in!\\
                !let y5 = 1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
        \end{center}  
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.\\
Of course, this term can be further optimised to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}
